---
title: "M05-Visualizing Data with Tables for MSDM CEP"
author: "Luis Ahumada"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    number-sections: true
    code-link: true
    embed-resources: true
editor: visual
execute: 
  freeze: auto
---

# Part 1: What I Learned from the Videos

## 1. Summary

After watching the videos, I learned that tables can be much more than just a basic way to showcase numbers and data. The `gt` package can help me make tables look much more organized and professional by adding titles, labels, colors, footnotes, and number formatting. I also learned that the data should be summarized and cleaned before create the final table. With `gtExtra` I learned that we can add small visuals directly inside a table which is nice because I thought that tables and charts had to be created separate when I would created them. By combing them, the data that I want to be present can become a lot easier to understand, rather than having things be all over the place.

::: callout-note
The key idea that I took away from the videos was that `gt` and `gExtra` packages can convert a basic table into something that is visually much easier to understand, and to organize as well.
:::

## 2. What I like about gt and gtExtra

What I liked most about the `gt` package was the amount of control it gives me to in terms of how the my table looks without making things too complicated. With this package, I can change column names, add subtitles, format numers, change colors, and include notes that can help me explain data better to my audience. These small perks can make a difference when building a table with data because I can make it much more professional and not so generic and confusing. What I liked about `gtExtra` was the fact that I can add more visual elements to my table like small charts and trends. This is useful because it give me the exact values and a quick idea of what is happening in the data, making it easier for me to explain it too. So, instead of choosing between table or chart, I can combine them using these packages.

## 3. How Do Tables Complement Charts?

Tables can complement charts by providing exact details that we may not be able to see on a chart as clearly. We can easily see patterns, trends, or differences in data when we are looking at a chart, and by also including a table, we can support the visual by showing the actual data behind it. An example would be that I might use a chart to show how sales have changed at the RKR over time, but management might want the actual numbers as well. So including a chart can show the trend of the RKR performance, and the actual numbers behind it allowing management to have a visual understanding and detailed one as well.

## 4. When I would prefer to use tables

I would prefer to use a table when the numerical values are more important than an overall pattern that can be represented with a visual. I can also use tables when comparing many measurements for a small number of categories. An example would be if I wanted to compare thee sales , and overall revenue of specific items on the RKR menu. I would use a table in this situation because it would allow me to see all the exact values in one place. Overall, I do think charts are important because they allow trends to be easily visualized, but tables are better used when the audience wants to see hard detailed numbers or information.

# Part 2: Creating the Table

## Analytics Objective:

For this section, I do not have the data that I will be receiving just yet so I will be using the `pizzaplace` data set to practice creating the table. For this table, I will use sales volume, average price, total revenue, revenue share, and monthly revenue trends. So, even though this data set is different than what I would use, I could still apply the same building approach when looking at Facebook engagement and RKR reservation metrics.

## Data Preparation:

First thing I will do is summarize the `pizzaplace` data by pizza type so that I could compare main categories. I calculated the number of pizzas sold, average price, total revenue, and each category's share of total revenue.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(gtExtras)
```

```{r}
#| label: pizza-summary
#| code-fold: true


pizza_summary <- pizzaplace |>
  group_by(type) |>
  summarise(
    pizzas_sold = n(),
    avg_price = mean(price),
    total_revenue = sum(price),
    .groups = "drop"
  ) |>
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )

pizza_summary
```

## Monthly Revenue Trend:

I will then summarize the data by month and pizza type so that I could see revenue changes throughout the year. These values are going to be used to create a small chart that shows the trend in the final table.

```{r}

#| label: monthly-trend
#| code-fold: true

monthly_revenue <- pizzaplace |>
  mutate(
    month = format(as.Date(date), "%m")
  ) |>
  group_by(type, month) |>
  summarise(
    revenue = sum(price),
    .groups = "drop"
  )

monthly_revenue

```

## Preparing the Monthly Trend:

I then grouped the monthly revenue values for each type of pizza so that I can use them to create a small trend chart in the final table.

```{r}

#| label: trend-data
#| code-fold: true

trend_data <- monthly_revenue |>
  arrange(type, month) |>
  group_by(type) |>
  summarise(
    monthly_trend = list(revenue),
    .groups = "drop"
  )

trend_data

```

## Combining the Summary Data:

I will then combine the descriptive stats with the monthly trend data so that the final table can show both values and small visual trend for each type of pizza.
