---
title: "M05-Visualizing Data with Tables for MSDM CEP"
author: "Luis Ahumada"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    number-sections: true
    code-link: true
    embed-resources: true
editor: visual
execute: 
  freeze: auto
---

# Part 1: What I Learned from the Videos

## 1. Summary

After watching the videos, I learned that tables can be much more than just a basic way to showcase numbers and data. The `gt` package can help me make tables look much more organized and professional by adding titles, labels, colors, footnotes, and number formatting. I also learned that the data should be summarized and cleaned before create the final table. With `gtExtra` I learned that we can add small visuals directly inside a table which is nice because I thought that tables and charts had to be created separate when I would created them. By combing them, the data that I want to be present can become a lot easier to understand, rather than having things be all over the place.

::: callout-note
The key idea that I took away from the videos was that `gt` and `gExtra` packages can convert a basic table into something that is visually much easier to understand, and to organize as well.
:::

## 2. What I like about gt and gtExtra

What I liked most about the `gt` package was the amount of control it gives me to in terms of how the my table looks without making things too complicated. With this package, I can change column names, add subtitles, format numers, change colors, and include notes that can help me explain data better to my audience. These small perks can make a difference when building a table with data because I can make it much more professional and not so generic and confusing. What I liked about `gtExtra` was the fact that I can add more visual elements to my table like small charts and trends. This is useful because it give me the exact values and a quick idea of what is happening in the data, making it easier for me to explain it too. So, instead of choosing between table or chart, I can combine them using these packages.

## 3. How Do Tables Complement Charts?

Tables can complement charts by providing exact details that we may not be able to see on a chart as clearly. We can easily see patterns, trends, or differences in data when we are looking at a chart, and by also including a table, we can support the visual by showing the actual data behind it. An example would be that I might use a chart to show how sales have changed at the RKR over time, but management might want the actual numbers as well. So including a chart can show the trend of the RKR performance, and the actual numbers behind it allowing management to have a visual understanding and detailed one as well.

## 4. When I would prefer to use tables

I would prefer to use a table when the numerical values are more important than an overall pattern that can be represented with a visual. I can also use tables when comparing many measurements for a small number of categories. An example would be if I wanted to compare thee sales , and overall revenue of specific items on the RKR menu. I would use a table in this situation because it would allow me to see all the exact values in one place. Overall, I do think charts are important because they allow trends to be easily visualized, but tables are better used when the audience wants to see hard detailed numbers or information.

# Part 2: Creating the Table

## Analytics Objective:

For this section, I do not have the data that I will be receiving just yet so I will be using the `pizzaplace` data set to practice creating the table. For this table, I will use sales volume, average price, total revenue, revenue share, and monthly revenue trends. So, even though this data set is different than what I would use, I could still apply the same building approach when looking at Facebook engagement and RKR reservation metrics.

## Data Preparation:

First thing I will do is summarize the `pizzaplace` data by pizza type so that I could compare main categories. I calculated the number of pizzas sold, average price, total revenue, and each category's share of total revenue.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(gtExtras)
```

```{r}
#| label: pizza-summary
#| code-fold: true


pizza_summary <- pizzaplace |>
  group_by(type) |>
  summarise(
    pizzas_sold = n(),
    avg_price = mean(price),
    total_revenue = sum(price),
    .groups = "drop"
  ) |>
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )

pizza_summary
```

## Monthly Revenue Trend:

I will then summarize the data by month and pizza type so that I could see revenue changes throughout the year. These values are going to be used to create a small chart that shows the trend in the final table.

```{r}

#| label: monthly-trend
#| code-fold: true

monthly_revenue <- pizzaplace |>
  mutate(
    month = format(as.Date(date), "%m")
  ) |>
  group_by(type, month) |>
  summarise(
    revenue = sum(price),
    .groups = "drop"
  )

monthly_revenue

```

## Preparing the Monthly Trend:

I then grouped the monthly revenue values for each type of pizza so that I can use them to create a small trend chart in the final table.

```{r}

#| label: trend-data
#| code-fold: true

trend_data <- monthly_revenue |>
  arrange(type, month) |>
  group_by(type) |>
  summarise(
    monthly_trend = list(revenue),
    .groups = "drop"
  )

trend_data

```

## Combining the Summary Data:

I will then combine the descriptive stats with the monthly trend data so that the final table can show both values and small visual trend for each type of pizza.

```{r}

#| label: combined-data
#| code-fold: true

pizza_table_data <- pizza_summary |>
  left_join(trend_data, by = "type")

pizza_table_data
```

## Creating the Final Table:

I will then use `gt` to convert the summarized data into a cleaner more professional table. I also included values and labels to make the table easier to read. You can see the result of the table below the chunk.

```{r}
#| label: final-gt-table
#| code-fold: true

pizza_table_data |>
  gt(rowname_col = "type") |>
  tab_stubhead(
    label = "Pizza Type"
  ) |>
  tab_header(
    title = "Pizza Sales Performance",
    subtitle = "Sales, pricing, revenue, and monthly trends by pizza type"
  ) |>
  cols_label(
    pizzas_sold = "Pizzas Sold",
    avg_price = "Average Price",
    total_revenue = "Total Revenue",
    revenue_share = "Revenue Share",
    monthly_trend = "Monthly Trend"
  )
```

## Formatting the Table:

I then will format the table to make it easier to read and more suitable for a presentation. I added labels, grouped related columns, and formatted numeric values.

```{r}
#| label: formatted-gt-table
#| code-fold: true

pizza_table_data |>
  gt(rowname_col = "type") |>
  tab_stubhead(
    label = "Pizza Type"
  ) |>
  tab_header(
    title = "Pizza Sales Performance",
    subtitle = "Sales, pricing, revenue, and monthly trends by pizza type"
  ) |>
  cols_label(
    pizzas_sold = "Pizzas Sold",
    avg_price = "Average Price",
    total_revenue = "Total Revenue",
    revenue_share = "Revenue Share",
    monthly_trend = "Monthly Trend"
  ) |>
  tab_spanner(
    label = "Sales Performance",
    columns = c(
      pizzas_sold,
      avg_price,
      total_revenue,
      revenue_share
    )
  ) |>
  fmt_number(
    columns = pizzas_sold,
    decimals = 0
  ) |>
  fmt_currency(
    columns = c(avg_price, total_revenue),
    currency = "USD",
    decimals = 2
  ) |>
  fmt_percent(
    columns = revenue_share,
    decimals = 2
  )
```

## Adding a Small Trend Chart:

```{r}
pizza_table_data |>
  gt(rowname_col = "type") |>
  tab_stubhead(
    label = "Pizza Type"
  ) |>
  tab_header(
    title = "Pizza Sales Performance",
    subtitle = "Sales, pricing, revenue, and monthly trends by pizza type"
  ) |>
  cols_label(
    pizzas_sold = "Pizzas Sold",
    avg_price = "Average Price",
    total_revenue = "Total Revenue",
    revenue_share = "Revenue Share",
    monthly_trend = "Monthly Trend"
  ) |>
  tab_spanner(
    label = "Sales Performance",
    columns = c(
      pizzas_sold,
      avg_price,
      total_revenue,
      revenue_share
    )
  ) |>
  fmt_number(
    columns = pizzas_sold,
    decimals = 0
  ) |>
  fmt_currency(
    columns = c(avg_price, total_revenue),
    currency = "USD",
    decimals = 2
  ) |>
  fmt_percent(
    columns = revenue_share,
    decimals = 2
  ) |>
  gt_plt_sparkline(
    column = monthly_trend,
    same_limit = TRUE,
    label = FALSE
  )
```

## Final Details:

I added color, a footnote, and a source note to make the table easier to read. The monthly sparklines are still in there so that the table shows the trends and values.

```{r}

#| label: final-pizza-table
#| code-fold: true

pizza_table_data |>
  gt(rowname_col = "type") |>
  tab_stubhead(
    label = "Pizza Type"
  ) |>
  tab_header(
    title = "Pizza Sales Performance",
    subtitle = "Sales, pricing, revenue, and monthly trends by pizza type"
  ) |>
  cols_label(
    pizzas_sold = "Pizzas Sold",
    avg_price = "Average Price",
    total_revenue = "Total Revenue",
    revenue_share = "Revenue Share",
    monthly_trend = "Monthly Trend"
  ) |>
  tab_spanner(
    label = "Sales Performance",
    columns = c(
      pizzas_sold,
      avg_price,
      total_revenue,
      revenue_share
    )
  ) |>
  fmt_number(
    columns = pizzas_sold,
    decimals = 0
  ) |>
  fmt_currency(
    columns = c(avg_price, total_revenue),
    currency = "USD",
    decimals = 2
  ) |>
  fmt_percent(
    columns = revenue_share,
    decimals = 2
  ) |>
  data_color(
    columns = total_revenue,
    method = "numeric",
    palette = "Blues"
  ) |>
  gt_plt_sparkline(
    column = monthly_trend,
    same_limit = TRUE,
    label = FALSE,
    palette = c(
      "black",
      "black",
      "purple",
      "green",
      "lightgrey"
    )
  ) |>
  tab_footnote(
    footnote = "Total revenue is the sum of the selling prices for all pizzas sold.",
    locations = cells_column_labels(
      columns = total_revenue
    )
  ) |>
  tab_source_note(
    source_note = "Source: pizzaplace dataset from the gt package."
  )
```

## Why This Table Is Worth Adding to MSDM Project:

This table is worth it because it would be super useful for what I am doing in my CEP. I will be able to keep many important measures in one place, making easy to compare performance, while the small trends chart can help me showcase how trends have changed over time. With the Facebook data that I will receive, I can compare content types, engagement, and reservation activity in a similar was and be useful for our written report and presentation slides.

# Appendix

## Project Links:

[Github Repository](https://github.com/lahumada63/Visualizing-Data-with-Tables-for-MSDM-CEP-Assignment.git)

Github Pages
